Illustrating splines using the Titanic dataset

### 2024-05-27 text/24/diamond-spline.Rmd
Moved r-datasets from pending to 24

```{r, echo=FALSE}
library(broom)
library(pspline)
library(readr)
library(rms)
library(tidyverse)
library(yaml)
f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
f1 <- "diamonds-data-dictionary.yaml"
dd <- read_yaml(paste0(f0, f1))
```

@@

```{r}
f2 <- "diamonds.txt"
raw_data <- read.fwf(file=paste0(f0, f2), widths=c(5,2,5,4,5))
di <- raw_data
names(di) <- c(
  "carat",
  "colour",
  "clarity",
  "certification",
  "price"
)
```

@@

```{r}
exchange_rate <- 1.72
inflation_rate <- 1.81
di$adjusted_price <- di$price*inflation_rate/exchange_rate
head(di)
```

@@

```{r}
summary(di$carat)
summary(di$price)
summary(di$adjusted_price)
count(di, colour)
count(di, clarity)
count(di, certification)
```

@@

```{r}
ggplot(data=di, aes(carat, adjusted_price)) +
  geom_point(pch=1) -> scatter1
scatter1
```

@@

```{r}
scatter1+geom_smooth(method="lm")
```

@@

```{r}
linear_fit <- lm(adjusted_price ~ carat, data=di)
linear_coefficients <- tidy(linear_fit)
linear_coefficients
```

@@

```{r}
spline_fit <- lm(adjusted_price ~ rcs(carat), data=di)
tidy(spline_fit)
```

@@

```{r}
carat_sequence <- data.frame(
  carat = seq(min(di$carat), max(di$carat), length=100))
carat_sequence$prediction <- 
  predict(spline_fit, newdata=carat_sequence)
scatter1 + geom_line(
  data=carat_sequence,
  aes(carat, prediction))
```

@@

```{r}
linear_plus_spline_fit <- lm(
  adjusted_price ~ carat + rcs(carat), 
  data=di)
anova(linear_fit, linear_plus_spline_fit)
```

>>>>

```{r initialization, echo=FALSE}
library(broom)
library(pspline)
library(readr)
library(rms)
library(tidyverse)
library(yaml)
f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
f1 <- "diamonds.yaml"
dd <- read_yaml(paste0(f0, f1))
```

@@

```{r read-diamonds}
f2 <- "diamonds.txt"
raw_data <- read.fwf(file=paste0(f0, f2), widths=c(5,2,5,4,5))
di <- raw_data
names(di) <- c(
  "carat",
  "colour",
  "clarity",
  "certification",
  "price"
)
```

@@

```{r adjustments}
exchange_rate <- 1.72
inflation_rate <- 1.81
di$adjusted_price <- di$price*inflation_rate/exchange_rate
head(di)
```

@@

```{r descriptives}
summary(di$carat)
summary(di$price)
summary(di$adjusted_price)
count(di, colour)
count(di, clarity)
count(di, certification)
```

@@

```{r plot}
ggplot(data=di, aes(carat, adjusted_price)) +
  geom_point(pch=1) -> scatter1
scatter1
```

@@

```{r linear-trend}
scatter1+geom_smooth(method="lm")
```

@@

```{r linear-fit}
linear_fit <- lm(adjusted_price ~ carat, data=di)
linear_coefficients <- tidy(linear_fit)
linear_coefficients
```

@@

```{r spline-fit}
spline_fit <- lm(adjusted_price ~ rcs(carat), data=di)
tidy(spline_fit)
```

@@

```{r spline-plot}
carat_sequence <- data.frame(
  carat = seq(min(di$carat), max(di$carat), length=100))
carat_sequence$prediction <- 
  predict(spline_fit, newdata=carat_sequence)
scatter1 + geom_line(
  data=carat_sequence,
  aes(carat, prediction))
```

@@

```{rlinear-plus-spline}
linear_plus_spline_fit <- lm(
  adjusted_price ~ carat + rcs(carat), 
  data=di)
anova(linear_fit, linear_plus_spline_fit)
```

### 2024-05-27 text/24/survival-lasso.Rmd
Moved r-datasets from pending to 24

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(broom)
library(glmnet)
library(glue)
library(pspline)
library(readr)
library(rms)
library(survival)
library(tidyverse)
library(yaml)
g0 <- "https://raw.githubusercontent.com/pmean/"
r0 <- "datasets/master/"
s1 <- "yaml-functions.R"
f1 <- "whas500.yaml"

@@

```{r}
vnames <- names(dd$vars)
f2 <- "whas500.dat"
w0 <- read_table(
  file=paste0(g0, r0, f2), 
  col_names=c(vnames),
  col_types="nnnnnnnnnnnnnnncccnnnn")
head(w0)
```

@@

```{r}
categorical <- 
  (vscales=="binary") |
  (vscales=="nominal") |
  (vscales=="ordinal")
for (i_vars in which(categorical)) {
  cat(glue("{vnames[i_vars]}: {vlabels[i_vars]}"), "\n")
  w0 %>%
    count(.data[[vnames[i_vars]]]) %>%
    data.frame %>%
    print
  cat("\n")
}
```

```{r}
continuous <- 
  (vscales=="interval") |
  (vscales=="ratio")
for (i_vars in which(continuous)) {
  cat(glue("{vnames[i_vars]}: {vlabels[i_vars]}"), "\n")
  print(summary(w0[vnames[i_vars]]))
  cat("\n")
}
```

@@

@@

```{r}
plot(Surv(w0$lenfol, w0$fstat))
```

@@

```{r}
simple_fit <- coxph(
  Surv(w0$lenfol, w0$fstat) ~
      cvd +
      afb + 
      sho + 
      chf + 
      av3 +
      miord +
      mitype,
    data=w0)
coef(simple_fit)
```

@@

```{r}
ymat <- cbind(w0$lenfol, w0$fstat)
dimnames(ymat)[2] <- list(c("time", "status"))
xmat <- as.matrix(w0[ , 8:14])
dimnames(xmat)[2] <- list(vnames[8:14])
lasso_fit <- glmnet(xmat, ymat, family="cox", standardize=FALSE)
plot(lasso_fit, xvar="lambda", label=TRUE)
```

>>>>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(glmnet)
library(glue)
library(readr)
library(survival)
library(tidyverse)
library(yaml)
g0 <- "https://raw.githubusercontent.com/pmean/"
r0 <- "datasets/master/"
s1 <- "yaml-functions.R"
f1 <- "whas500.yaml"

@@

```{r, echo=FALSE}
vnames <- names(dd$vars)
# for some reason, R wants to read in 23 
# columns rather than 22. The last column 
# is all blanks and will be tossed out.
#
# Dates are always a bit messy and I do
# not need them for this analysis, so I
# will keep things simple and read them 
# in as strings.
f2 <- "whas500.dat"
w0 <- read_table(
  file=paste0(g0, r0, f2), 
  col_names=vnames,
  col_types="nnnnnnnnnnnnnnncccnnnn")
w1 <- data.frame(w0)
head(w1)
```

@@

```{r, echo=FALSE}
categorical <- 
  (vscales=="binary") |
  (vscales=="nominal") |
  (vscales=="ordinal")
for (i_vars in which(categorical)) {
  cat(glue("{vnames[i_vars]}: {vlabels[i_vars]}"), "\n")
  w1 %>%
    count(.data[[vnames[i_vars]]]) %>%
    data.frame %>%
    print
  cat("\n")
}
```

```{r, echo=FALSE}
continuous <- 
  (vscales=="interval") |
  (vscales=="ratio")
for (i_vars in which(continuous)) {
  cat(glue("{vnames[i_vars]}: {vlabels[i_vars]}"), "\n")
  print(summary(w1[vnames[i_vars]]))
  cat("\n")
}
```

@@

I am going to show the R code from this point onward so you can see the actual programming required to fit a lasso model.

@@

```{r}
plot(Surv(w1$lenfol, w1$fstat))
```

@@

```{r}
simple_fit <- coxph(
  Surv(w1$lenfol, w1$fstat) ~
      cvd +
      afb + 
      sho + 
      chf + 
      av3 +
      miord +
      mitype,
    data=w1)
coef(simple_fit)
```

@@

```{r}
ymat <- cbind(w1$lenfol, w1$fstat)
dimnames(ymat)[2] <- list(c("time", "status"))
xmat <- as.matrix(w1[ , 8:14])
dimnames(xmat)[2] <- list(vnames[8:14])
lasso_fit <- glmnet(xmat, ymat, family="cox", standardize=FALSE)
plot(lasso_fit, xvar="lambda", label=TRUE)
```

### 2024-05-27 text/24/titanic-imputation.Rmd
Moved r-datasets from pending to 24

Splines provide a useful way to model relationships that are more complex than a simple linear relationship. They work in a variety of regression models. Here is an illustration of how to use a spline in a logistic regression model with data from survival of passengers on the Titanic.

@@

```{r, echo=FALSE}
f2 <- "titanic_v00.txt"
raw_data <- 
  data.frame(
    read_tsv(
      file=paste0(g0, r0, f2),
      col_types="ccncn"))
# Re-arranging columns just for convenient display
o <-  c(1, 3, 2, 4, 5)
ti0 <- select(raw_data, o)
vnames <- vnames[o]
head(ti0)
```

>>>>

Multiple imputation is a way to properly account for missing values without causing bias. Simpler forms of imputation, such as replacing missing values with the mean of the non-missing values, can produce serious problems. You might think that ignoring the missingness and relying just on records with complete data for all key variables would be acceptable, but this also can produce serious problems. I want to illustrate a simple example of multiple imputation using data on mortality from the Titanic.

@@

```{r, echo=FALSE}
f2 <- "titanic.txt"
raw_data <- 
  data.frame(
    read_tsv(
      file=paste0(g0, r0, f2),
      col_types="ccncn"))
# Re-arranging columns just for convenient display
o <-  c(1, 3, 2, 4, 5)
ti0 <- select(raw_data, o)
vnames <- vnames[o]
head(ti0)
```

### 2024-05-27 text/24/titanic-spline.Rmd
Moved r-datasets from pending to 24

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(broom)
library(pspline)
library(readr)
library(rms)
library(tidyverse)
library(yaml)
f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
f1 <- "titanic-data-dictionary.yaml"
dd <- read_yaml(paste0(f0, f1))
```

@@

```{r}
f2 <- "titanic_v00.txt"
ti0 <- 
  data.frame(
    read_tsv(
      file=paste0(f0, f2),
      col_types="ccncn"))
names(ti0) <- tolower(names(ti0))
head(ti0)
```

>>>>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(broom)
library(pspline)
library(readr)
library(rms)
library(tidyverse)
library(yaml)
f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
f1 <- "titanic.yaml"
dd <- read_yaml(paste0(f0, f1))
```

@@

```{r}
f2 <- "titanic.txt"
ti0 <- 
  data.frame(
    read_tsv(
      file=paste0(f0, f2),
      col_types="ccncn"))
names(ti0) <- tolower(names(ti0))
head(ti0)
```
